---
title: "IRT"
author: "Peng-Ting Kuo"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Setup

```{asis Problems}
There are a few things that worth noticing:  
1. I recoded the symbolism indicator as it shows negative loading in single factor analysis (I recoded 2(deified) to 0 (plain) and 0 to 2). Also, symbolism only has 1400+ data with labeled as 3 (ceremonial).
2. Sovereign seems to be a single vector by itself
3. I also recoded assembly and election since they are related concept. The new indicator called representation has six labels
```

# Preprocessing

```{r setup, warning=FALSE}
suppressPackageStartupMessages({
    library(haven)
    library(ggplot2)
    library(tidyverse)
    library(mirt)
    library(psych)
    library(scales)
    library(MCMCpack)
    select <- dplyr::select
})

# mirt parallel compute
mirtCluster(4)

# Load datasets
plt_gemini <- read_csv("../data/plt_constraints.csv", show_col_types = F)
# region_broad has the least missingness (n = 16090)
raw_data <- read_csv("../data/plt_dataset.csv", show_col_types = F) %>%
    select(id, region_broad) %>%
    rename(region = region_broad)

plt_gemini <- left_join(plt_gemini, raw_data, by = "id")

# data aggregation
plt_gemini <- plt_gemini %>%
    mutate(
        entry_4_prediction = case_when(
            entry_prediction %in% c(0, 1) ~ 0,
            entry_prediction == 4 ~ 1,
            entry_prediction %in% c(2, 3, 5, 7, 8) ~ 2,
            entry_prediction %in% c(6, 9) ~ 3
        ),
        exit_4_prediction = case_when(
            exit_prediction %in% c(3, 4, 7, 8, 9, 10, 11, 12, 13) ~ 0,
            exit_prediction %in% c(5, 6) ~ 1,
            exit_prediction == 0 ~ 2,
            exit_prediction %in% c(1, 2) ~ 3
        ),
        checks_prediction = checks_local_prediction + checks_military_prediction +
            checks_judiciary_prediction + checks_council_prediction +
            checks_bureaucracy_prediction + checks_aristocracy_prediction +
            checks_assembly_prediction + checks_bourgeoisie_prediction +
            checks_clergy_prediction,
        elections_raw_prediction = elections_prediction,
        elections_prediction = if_else(assembly_prediction == 2,
                                       elections_prediction,
                                       NA_real_),
        # 0 - no
        # 1 - council, legislative with no election
        # 2 - assembly with non-competitive election
        # 3 - assembly with election and popular asembly
        representation = case_when(
            assembly_prediction == 0 ~ 0,
            assembly_prediction == 1 ~ 1,
            assembly_prediction == 2 & elections_prediction == 0 ~ 1,
            assembly_prediction == 2 & elections_prediction == 1 ~ 2,
            assembly_prediction == 2 & elections_prediction == 2 ~ 3,
            assembly_prediction == 3 ~ 3),
        # symbolism recode
        symbolism_recode = case_when(
            symbolism_prediction == 2 ~ 0,
            symbolism_prediction == 1 ~ 1,
            symbolism_prediction == 0 ~ 2,
            symbolism_prediction == 3 ~ 3,
        ),
        tenure = (leader_last_year - leader_first_year + 1),
        tenure_log = log(tenure),
        across(c(exit_prediction, entry_prediction), ~na_if(., 99)))

plt_gemini %>%
    group_by(polity_name) %>%
    mutate(win_span = leader_last_year - lag(leader_first_year, 2)) %>%
    ungroup() %>%
    summarise(across(win_span, list(p10 = ~quantile(.x, .1, na.rm = TRUE),
                                    med = ~median(.x, na.rm = TRUE),
                                    p90 = ~quantile(.x, .9, na.rm = TRUE))))
```

### Tenure

```{r}
# tenure classification
plt_gemini <- plt_gemini %>%
    drop_na(leader_first_year, leader_last_year) %>%
    arrange(polity_name, leader_first_year, leader_last_year) %>%
    group_by(polity_name) %>%
    mutate(
        gap   = leader_first_year - lag(leader_last_year),
        spell = cumsum(is.na(gap) | gap > 500)
    ) %>%
    group_by(polity_name, spell) %>%
    mutate(
        tenure_w1          = tenure,
        tenure_trailing_w3 = pmax(tenure, lag(tenure, 1), lag(tenure, 2), na.rm = TRUE),
        tenure_centered_w3 = pmax(lag(tenure, 1), tenure, lead(tenure, 1), na.rm = TRUE),
        tenure_leading_w3  = pmax(tenure, lead(tenure, 1), lead(tenure, 2), na.rm = TRUE),
        tenure_trailing_w5 = pmax(tenure, lag(tenure, 1), lag(tenure, 2),
                                  lag(tenure, 3), lag(tenure, 4), na.rm = TRUE),
        tenure_centered_w5 = pmax(lag(tenure, 2), lag(tenure, 1), tenure,
                                  lead(tenure, 1), lead(tenure, 2), na.rm = TRUE),
        tenure_leading_w5  = pmax(tenure, lead(tenure, 1), lead(tenure, 2),
                                  lead(tenure, 3), lead(tenure, 4), na.rm = TRUE),
        
        # same window as lexical_lw3; 2 = <8 years (most constrained), 0 = 16+
        tenure_3_prediction_w3 = case_when(
            tenure_leading_w3 <= 8 ~ 2,
            tenure_leading_w3 <= 15 ~ 1,
            tenure_leading_w3 > 15 ~ 0
        )
    ) %>%
    ungroup()

plt_gemini %>%
    group_by(polity_name, spell) %>%
    mutate(win_span = leader_last_year - lag(leader_first_year, 2)) %>%
    ungroup() %>%
    summarise(across(win_span, list(p10 = ~quantile(.x, .1, na.rm = TRUE),
                                    med = ~median(.x, na.rm = TRUE),
                                    p90 = ~quantile(.x, .9, na.rm = TRUE),
                                    max = ~max(.x, na.rm = TRUE))))

# Calender approach
plt_gemini <- plt_gemini %>%
    group_by(polity_name) %>%
    mutate(
        tenure_cal25 = sapply(leader_last_year, function(t)
            max(tenure[leader_last_year <= t & leader_last_year > t - 25], na.rm = TRUE))
    ) %>%
    ungroup()


span <- plt_gemini %>%
    arrange(polity_name, leader_first_year) %>%
    group_by(polity_name) %>%
    mutate(win_span = leader_last_year - lag(leader_first_year, 2)) %>%
    ungroup() %>%
    filter(!is.na(win_span), win_span > 0)

# 1. Distribution of tenure window
ggplot(span, aes(win_span)) +
    geom_histogram(bins = 50) +
    scale_x_log10(breaks = c(1, 2, 5, 12, 25, 71, 200)) +
    geom_vline(xintercept = 12, linetype = "dashed") +
    labs(x = "Total year of three leaders (log)", y = "Leader Number") +
    theme_minimal()

# Distribution across time
span %>%
    mutate(era = floor(leader_last_year / 100) * 100) %>%
    group_by(era) %>%
    summarise(n = n(),
              p10 = quantile(win_span, .1),
              med = median(win_span),
              p90 = quantile(win_span, .9)) %>%
    filter(n >= 30) %>%
    ggplot(aes(era, med)) +
    geom_ribbon(aes(ymin = p10, ymax = p90), alpha = .2) +
    geom_line() +
    scale_y_log10() +
    labs(x = "Year", y = "Window Length (log)") +
    theme_minimal()

# Something's wrong in the polity and leader year
plt_gemini %>%
    filter(polity_name %in% c("Mari", "Anhalt Freistaat", "Timbuktu")) %>%
    arrange(polity_name, leader_first_year) %>%
    select(polity_name, leader_name, leader_first_year, leader_last_year) %>%
    print(n = 100)

# Same polity has BCE and CE
plt_gemini %>%
    group_by(polity_name) %>%
    summarise(n = n(),
              n_bce = sum(leader_first_year < 0, na.rm = TRUE),
              n_ce  = sum(leader_first_year > 0, na.rm = TRUE)) %>%
    filter(n_bce > 0, n_ce > 0) %>%
    arrange(n_bce) %>%
    print(n = 50)

plt_gemini <- plt_gemini %>%
    mutate(name_style = if_else(str_detect(leader_name, "^[a-z]|[()]"), "messy", "clean"))

count(plt_gemini, name_style)

plt_gemini %>% filter(gap > 500) %>% count(polity_name, sort = TRUE)
plt_gemini %>% count(polity_name, spell) %>% filter(n == 1) %>% nrow()
```

### Lexical check

```{r}
plt_gemini %>%
    summarise(across(starts_with("tenure_") & !matches("^tenure$"),
                     list(n = ~sum(!is.na(.x)), lvl1 = ~mean(.x > 15, na.rm = TRUE)))) %>%
    pivot_longer(everything(), names_to = c("spec", ".value"), names_pattern = "(.*)_(n|lvl1)$")


# validity
py <- plt_gemini %>%
    mutate(year = map2(leader_first_year, leader_last_year, seq)) %>%
    unnest(year) %>%
    group_by(polity_name, year) %>%
    summarise(across(starts_with("tenure_") & !matches("log"),
                     ~ if (all(is.na(.x))) NA_real_ else max(.x, na.rm = TRUE)),
              .groups = "drop")

py %>%
    summarise(across(starts_with("tenure_"), ~mean(.x > 15, na.rm = TRUE)))


# Why 15?
thresh <- map_dfr(5:40, function(k) {
    py %>%
        summarise(across(c(tenure_w1, tenure_trailing_w3, tenure_trailing_w5),
                         ~mean(.x > k, na.rm = TRUE))) %>%
        mutate(threshold = k)
})

thresh %>%
    pivot_longer(-threshold, names_to = "spec", values_to = "share") %>%
    ggplot(aes(threshold, share, colour = spec)) +
    geom_line() +
    geom_vline(xintercept = 15, linetype = "dashed") +
    labs(y = "share at lexical level 1") +
    theme_minimal()
```

### Checks & Representation (Assembly + Election)

```{r}
# Distribution of checks
ggplot(plt_gemini, aes(x=checks_prediction)) +
    geom_bar(fill = "steelblue", width = 0.7) +
    labs(title = " Aggregated Checks Distribution", x = "Labels", y = "Values") +
    scale_x_continuous(breaks = 0:9) +
    geom_text(aes(label = after_stat(count)),
              stat = "count",
              vjust = -0.3,
              size = 4) + 
    theme_bw()

# Distribution of representation
ggplot(plt_gemini, aes(x=factor(representation))) +
    geom_bar(fill = "steelblue", width = 0.7) +
    labs(title = "Representation Distribution", x = "Labels", y = "Values") +
    scale_x_discrete(labels = c("0" = "No Assembly",
                                "1" = "Council",
                                "2" = "Legislature, no elections",
                                "3" = "Non-competitive Legislature",
                                "4" = "Competitive Legislature",
                                "5" = "Popular assembly")) +
    geom_text(aes(label = after_stat(count)),
              stat = "count",
              hjust = -0.3,
              size = 4) +
    theme_bw() +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) 

# checks_prediction is sparse after 7 (8: n = 189 (0.14%), 9: n = 33 (0.02%))
# combine 7/8/9 to "7+" (1,274 in total)
plt_gemini <- plt_gemini %>%
    mutate(checks_prediction =  pmin(checks_prediction, 7),
           era = cut(leader_first_year, c(-Inf, 0, 1500, 1800, Inf),
                     labels = c("BC","0-1500","1500-1800","1800-"))) %>%
    mutate(checks_3_prediction = case_when(
        checks_prediction == 0 ~ 0,
        checks_prediction %in% c(1, 2, 3) ~ 1,
        checks_prediction > 3 ~ 2
    ))
```

### Entry & Exit

```{r}
# See the facial validity looks like for entry indicator with four lables
entry_table <- plt_gemini %>%
    mutate(
        total_score = rowSums(select(., sovereign_prediction, constitution_prediction,  
                                     federalism_prediction, 
                                     checks_prediction, collegiality_prediction,
                                     petition_prediction, assembly_prediction,
                                     elections_prediction), na.rm = TRUE)
    ) %>%
    group_by(entry_4_prediction, era) %>%
    summarise(
        n = n(),
        mean_score = mean(total_score, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    filter(!is.na(entry_4_prediction) & !is.na(era)) %>%
    mutate(display_val = paste0(round(mean_score, 2), " (n=", n, ")")) %>%
    select(entry_4_prediction, era, display_val) %>%
    pivot_wider(names_from = era, values_from = display_val)

print("---Entry Indicator Table---")
print(entry_table)

# exit
exit_table <- plt_gemini %>%
    mutate(
        total_score = rowSums(select(., sovereign_prediction, constitution_prediction, 
                                     federalism_prediction, checks_prediction, 
                                     collegiality_prediction, petition_prediction, 
                                     assembly_prediction, elections_prediction), na.rm = TRUE)
    ) %>%
    group_by(exit_4_prediction, era) %>%
    summarise(
        n = n(),
        mean_score = mean(total_score, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    filter(!is.na(exit_4_prediction) & !is.na(era)) %>%
    mutate(display_val = paste0(round(mean_score, 2), " (n=", n, ")")) %>%
    select(exit_4_prediction, era, display_val) %>%
    pivot_wider(names_from = era, values_from = display_val)

print("---Exit Indicator Table---")
print(exit_table)
```

### Quasi-lexical scale

```{asis Lexical}
Higher scores supersede lower ones; Tenure > 15 caps the score at 1 (two specs: own tenure, leading 3-leader window).
1 none; 2 code of law (or petitions); 3 constitution (or ceremonial leader / federalism);
4 advisory body; 5 checks (or collegial / entry by appointment); 6 legislative body;
7 elective leader (entry = 9); 8 regular exit (elective leader, exit 0/1/2).
```

```{r Quasi-lexical}
# Drop rows with unknown entry/exit (99 -> NA in setup). Done AFTER the tenure
# windows are built so lead()/lag() still see every leader in the polity.
plt_gemini %>%
    mutate(dropped = is.na(entry_prediction) | is.na(exit_prediction)) %>%
    group_by(era) %>%
    summarise(n = n(), dropped = sum(dropped), share = round(dropped / n, 3))

lex <- plt_gemini %>%
    drop_na(entry_prediction, exit_prediction)

# Criterion met for each level (1 = met). "Assumed if" clauses qualify a row
# for that level directly, since higher scores supersede lower ones.
lex <- lex %>%
    mutate(
        lex2_code_of_law  = constitution_prediction >= 1 | petition_prediction == 1,
        lex3_constitution = constitution_prediction == 2 | symbolism_prediction == 3 |
                            federalism_prediction == 1,
        # council or non-competitive legislature; cumulative, so any higher
        # representative body also counts
        lex4_advisory     = assembly_prediction >= 1| representation == 2,
        lex5_checks       = checks_3_prediction >= 1 | collegiality_prediction == 1 |
                            entry_4_prediction %in% 2,
        lex6_legislative  = assembly_prediction == 3 |
                            (assembly_prediction == 2 & elections_prediction %in% 2),
        lex7_elective     = entry_prediction == 9,
        lex8_regular_exit = lex7_elective & exit_prediction %in% c(0, 1, 2),
        across(matches("^lex[2-8]_"), as.integer)
    )

lex_items <- grep("^lex[2-8]_", names(lex), value = TRUE)

# Highest level whose criterion is met; Tenure > 15 caps the score at 1
lexical_score <- function(data, tenure_col = NULL) {
    score <- rep(1L, nrow(data))
    for (k in 2:8) score[data[[lex_items[k - 1]]] %in% 1] <- k
    if (!is.null(tenure_col)) score[data[[tenure_col]] > 15] <- 1L
    score
}

lex <- lex %>%
    mutate(lexical_raw = lexical_score(lex),
           lexical_w1  = lexical_score(lex, "tenure_w1"),
           lexical_lw3 = lexical_score(lex, "tenure_leading_w3"))

# Level distribution: no cap vs. hard cap on own tenure vs. leading 3-leader window
lex %>%
    select(lexical_raw, lexical_w1, lexical_lw3) %>%
    pivot_longer(everything(), names_to = "spec", values_to = "level") %>%
    count(spec, level) %>%
    group_by(spec) %>%
    mutate(share = round(n / sum(n), 3)) %>%
    ungroup() %>%
    pivot_wider(id_cols = level, names_from = spec, values_from = share)

# Empirical demandingness: share meeting each criterion (should fall with level)
lex %>%
    summarise(across(all_of(lex_items), ~round(mean(.x), 3))) %>%
    pivot_longer(everything(), names_to = "level", values_to = "share_met")

# Exceptions: among rows meeting the higher criterion (column), share NOT
# meeting the lower one (row). Level 8 is conditional on 7 by construction.
guttman_items <- lex_items[1:6]
m <- as.matrix(lex[guttman_items])
exceptions <- matrix(NA_real_, 6, 6, dimnames = list(guttman_items, guttman_items))
for (i in 1:5) for (j in (i + 1):6) {
    exceptions[i, j] <- round(sum(m[, j] == 1 & m[, i] == 0) / sum(m[, j] == 1), 3)
}
exceptions

# Scalability of an ordering (easiest -> most demanding):
#   CR = 1 - Guttman errors / (n * items), errors against the ideal pattern
#        implied by the quasi-lexical score
#   H  = 1 - observed / expected pairwise errors (Loevinger's H, theoretical order)
scalability <- function(m, order) {
    m <- m[, order, drop = FALSE]
    K <- ncol(m)
    top <- apply(m, 1, function(r) if (any(r == 1)) max(which(r == 1)) else 0L)
    ideal <- outer(top, seq_len(K), ">=") * 1L
    p <- colMeans(m)
    obs <- 0; expd <- 0
    for (i in 1:(K - 1)) for (j in (i + 1):K) {
        obs  <- obs + sum(m[, j] == 1 & m[, i] == 0)
        expd <- expd + nrow(m) * p[j] * (1 - p[i])
    }
    tibble(CR = round(1 - sum(m != ideal) / length(m), 3),
           H  = round(1 - obs / expd, 3))
}

# Robustness items:
#   lex4_strict: advisory body as written (council or non-competitive legislature),
#                NOT nested in level 6, so lex4/lex6 pairs are no longer errors by definition
#   lex5_nobody: checks excluding checks_assembly / checks_council, which overlap
#                mechanically with the representation levels 4 and 6
lex <- lex %>%
    mutate(
        lex4_strict = as.integer(assembly_prediction == 1 |
                                 (assembly_prediction == 2 & elections_prediction %in% c(0, 1))),
        lex5_nobody = as.integer(
            (checks_prediction - checks_assembly_prediction - checks_council_prediction) >= 1 |
            collegiality_prediction == 1 | entry_4_prediction %in% 2)
    )

codings <- list(
    cumulative        = guttman_items,
    as_written        = replace(guttman_items, 3, "lex4_strict"),
    checks_no_bodies  = replace(guttman_items, 4, "lex5_nobody"),
    as_written_nobody = replace(guttman_items, 3:4, c("lex4_strict", "lex5_nobody"))
)

```

### Overall

```{r}
# missingness (we only allow entry and exit indicator to have NA)
miss_table <- plt_gemini %>%
    group_by(region, era) %>% 
    summarise(
        n = n(),
        miss_entry_rate = mean(is.na(entry_4_prediction)), 
        miss_exit_rate = mean(is.na(exit_4_prediction)),
        .groups = "drop"
    ) %>%
    filter(!is.na(region) & !is.na(era)) %>%
    mutate(display_val = paste0(
        percent(miss_entry_rate, accuracy = 0.1), " / ", 
        percent(miss_exit_rate, accuracy = 0.1), " (n=", n, ")"
    )) %>%
    select(region, era, display_val) %>%
    pivot_wider(names_from = era, values_from = display_val)

print("--- Combined Missingness Table (Entry % / Exit % (n=Total)) ---")
print(miss_table)

# Seems like there's a chronological missingness


# total score
summary_df <- plt_gemini %>%
    mutate(
        total_score = rowSums(select(., sovereign_prediction, constitution_prediction, 
                                     federalism_prediction, checks_prediction, 
                                     collegiality_prediction, petition_prediction, 
                                     assembly_prediction, elections_prediction), 
                              na.rm = TRUE)) %>%
    group_by(era, region) %>%
    summarise(
        n = n(),
        mean_score = mean(total_score, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    filter(!is.na(era) & !is.na(region))

table_df <- summary_df %>%
    mutate(display_val = paste0(round(mean_score, 2), " (n=", n, ")")) %>%
    select(region, era, display_val) %>%
    pivot_wider(names_from = era, values_from = display_val)

print("---Total score table---")
print(table_df)


robust_tbl <- plt_gemini %>%
    summarise(across(starts_with("tenure_"),
                     list(mean = ~mean(.x, na.rm = TRUE),
                          med  = ~median(.x, na.rm = TRUE),
                          lvl1 = ~mean(.x > 15, na.rm = TRUE)))) %>%
    pivot_longer(everything(),
                 names_to = c("spec", ".value"),
                 names_pattern = "tenure_(.*)_(mean|med|lvl1)$") %>%
    separate(spec, c("alignment", "width"), sep = "_w", fill = "left") %>%
    mutate(alignment = replace_na(alignment, "(none)"),
           width = as.integer(width)) %>%
    arrange(width, alignment)

print(robust_tbl)

robust_tbl %>%
    group_by(alignment) %>%
    summarise(ok = !is.unsorted(lvl1))
```

# PCA

```{r PCA}
# Version A: all 19 indicators
items_A <- c(
    "constitution_prediction",
    "sovereign_prediction",
    "federalism_prediction",
    "collegiality_prediction",
    "petition_prediction",
    "assembly_prediction",
    "elections_prediction",
    "symbolism_prediction",
    "entry_4_prediction",
    "exit_4_prediction",
    "checks_local_prediction",
    "checks_military_prediction",
    "checks_judiciary_prediction",
    "checks_council_prediction",
    "checks_bureaucracy_prediction",
    "checks_aristocracy_prediction",
    "checks_assembly_prediction",
    "checks_bourgeoisie_prediction",
    "checks_clergy_prediction"
)

# Version B: Checks_prediction aggregation
items_B <- c(
    "constitution_prediction",
    "sovereign_prediction",
    "federalism_prediction",
    "collegiality_prediction",
    "petition_prediction",
    "representation",
    # "assembly_prediction",
    # "elections_prediction",
    "symbolism_recode",
    "entry_4_prediction",
    "exit_4_prediction",
    "checks_prediction",
    "tenure_3_prediction_w3"
)

# Version C: exclude sovereign
items_C <- c(
    "constitution_prediction",
    "federalism_prediction",
    "collegiality_prediction",
    "petition_prediction",
    "representation",
    "symbolism_recode",
    "entry_4_prediction",
    "exit_4_prediction",
    "checks_prediction"
)

data_items_B <- plt_gemini[items_B]
lapply(data_items_B, function(x) round(prop.table(table(x, useNA = "ifany")), 3))
lapply(data_items_B, function(x) round(mean(x, na.rm = T), 3))
# missingness
plt_gemini %>% mutate(cc = complete.cases(data_items_B)) %>%
    group_by(era) %>% summarise(complete_rate = mean(cc))

fit_pc <- princomp(na.omit(data_items_B), cor = TRUE)
print(fit_pc)

# loadings
fit_pc$loadings

# screeplot
screeplot(fit_pc, type ='line')

```

# Factor Analysis

```{r FA}
# single factor
fa1 <- fa(data_items_B, nfactors = 1, fm = "pa")
print(fa1)

fit_fa <- fa(r = data_items_B,
             nfactors = 3,
             scores = "regression",
             rotate = "oblimin") # This is the default method

pc <- psych::polychoric(data_items_B, global = FALSE)

n_eff <- sum(complete.cases(data_items_B))

fa1_poly <- fa(r = pc$rho, nfactors = 1, n.obs = n_eff, fm = "pa")
fa3_poly <- fa(r = pc$rho, nfactors = 3, n.obs = n_eff, fm = "pa", rotate = "oblimin")

# How many factors do we need?
fa.parallel(data_items_B, fa = "fa", cor = "poly")


print(fa1_poly)

print(fa3_poly)
```

```{r Test}
psych::omega(data_items_B, nfactors = 3, poly = TRUE, fm = "pa")
```

# IRT Model

### MCMCpack (would take 5-7 hours)

```{r MCMCpack}
sub <- data_items_B[sample(nrow(data_items_B), 5000), ]
system.time(
    test <- MCMCordfactanal(as.matrix(sub), factors = 1,
                            lambda.constraints = list(representation = list(2, "+")),
                            burnin = 500, mcmc = 1000, thin = 1,
                            store.lambda = TRUE, store.scores = TRUE,
                            tune = 1.2, verbose = 500)
)
```

### Using mirt since emIRT is not applicable this time

```{r IRT}
#===============================================================================
# IRT: item sets
#===============================================================================

#-------------------------------------------------------------------------------
# helper: item matrix & itemtype. Mixed with "2PL" and "graded"
# recoded to K-1
#-------------------------------------------------------------------------------
make_irt_data <- function(df, items) {
    dat <- df %>% select(all_of(items)) %>% as.data.frame()
    
    for (v in items) {
        x <- dat[[v]]
        lv <- sort(unique(x[!is.na(x)]))
        cat(sprintf("%-32s n_cat=%d  levels=%s  NA=%d (%.1f%%)\n",
                    v, length(lv), paste(lv, collapse = ","),
                    sum(is.na(x)), 100 * mean(is.na(x))))
        dat[[v]] <- match(x, lv) - 1L   # recoded to 0, 1 ... K-1
    }
    dat
}

get_itemtype <- function(dat) {
    vapply(dat, function(x) {
        k <- length(unique(x[!is.na(x)]))
        if (k <= 2) "2PL" else "graded"
    }, character(1))
}

cat("\n===== Version A items =====\n")
dat_A <- make_irt_data(plt_gemini, items_A)
type_A <- get_itemtype(dat_A)

cat("\n===== Version B items =====\n")
dat_B <- make_irt_data(plt_gemini, items_B)
type_B <- get_itemtype(dat_B)

print(type_A)

dat_C  <- make_irt_data(plt_gemini, items_C)
type_C <- get_itemtype(dat_C)


set.seed(42)

mod_A <- mirt(dat_A, model = 1, itemtype = type_A,
              method = "EM", SE = TRUE,
              technical = list(NCYCLES = 2000))

mod_B <- mirt(dat_B, model = 1, itemtype = type_B,
              method = "EM", SE = TRUE,
              technical = list(NCYCLES = 2000))

m1 <- mirt(dat_C, 1, itemtype = type_C, method = "EM", technical = list(NCYCLES = 2000))
m2 <- mirt(dat_B, 2, itemtype = type_B, method = "EM", technical = list(NCYCLES = 2000))
m3 <- mirt(dat_B, 3, itemtype = type_B, method = "EM", technical = list(NCYCLES = 2000))

anova(m1, m2)
anova(m2, m3)

summary(m2, rotate = "oblimin")
summary(m3, rotate = "oblimin")

saveRDS(mod_A, "./mod_A_unidim.rds")
saveRDS(mod_B, "./mod_B_unidim.rds")

#===============================================================================
# Evaluation
#===============================================================================
coef_A <- coef(mod_A, simplify = TRUE, IRTpars = TRUE)$items
coef_B <- coef(mod_B, simplify = TRUE, IRTpars = TRUE)$items

round(coef_A, 3)
round(coef_B, 3)


coef_A %>%
    as.data.frame() %>%
    rownames_to_column("item") %>%
    select(item, a) %>%
    arrange(desc(a)) %>%
    print()

summary(mod_A)
summary(mod_B)

plt_gemini %>%
    filter((polity_name == "Soviet Union") & (leader_first_year >= 1924) & (leader_last_year <= 1953)) %>% View()
```

## Facial Validity

```{r Validity test}
# mirtCluster(remove = TRUE)
th1 <- fscores(m1, method = "EAP", full.scores.SE = TRUE)

plt_scored <- plt_gemini %>%
    mutate(theta = th1[, 1], theta_se = th1[, 2],
           n_miss = rowSums(is.na(dat_B)))
# Era
plt_scored %>%
    group_by(era) %>%
    slice_max(theta, n = 10) %>%
    select(era, polity_name, leader_name, leader_first_year, theta, theta_se) %>%
    arrange(era, desc(theta)) %>%
    print(n = 50)

plt_scored %>%
    group_by(era) %>%
    slice_min(theta, n = 10) %>%
    select(era, polity_name, leader_name, leader_first_year, theta, theta_se) %>%
    print(n = 50)

# Polity (It seems like it depends on the length of that polity as well)
plt_scored %>%
    group_by(polity_name) %>%
    filter(n() >= 30) %>%
    summarise(n = n(), theta = mean(theta, na.rm = TRUE)) %>%
    arrange(desc(theta)) %>%
    print(n = 20)


# Known polities
source("classic_cases.R")
# key_cases <- c("Roman Republic","Athens","Venice","England",
#                "Russia","Prussia","Ottoman Empire","China",
#                "Soviet Union","Germany","Japan","France", "United States Of America")
plt_scored %>%
    add_case() %>%
    group_by(case) %>%
    summarise(n          = n(),
              n_leaders  = n_distinct(leader_name),
              theta_mean = mean(theta, na.rm = TRUE),
              theta_sd   = sd(theta, na.rm = TRUE),
              theta_sem  = theta_sd / sqrt(sum(!is.na(theta))),
              eap_se     = mean(theta_se, na.rm = TRUE),
              .groups    = "drop") %>%
    arrange(desc(theta_mean)) %>% print(n = Inf)

ggplot(plt_scored, aes(leader_first_year, theta)) +
    geom_point(alpha = 0.03) +
    geom_smooth(method = "gam") +
    labs(x = "Year", y = "Executive constraints")

th2 <- fscores(m2, method = "EAP") 
plt2 <- plt_scored %>% mutate(f1 = th2[,1], f2 = th2[,2])


case_pts <- plt2 %>% add_case()     

case_ctr <- case_pts %>%
    group_by(case) %>%
    summarise(f1 = mean(f1, na.rm = TRUE),
              f2 = mean(f2, na.rm = TRUE),
              n  = n(), .groups = "drop")   

ggplot(plt2, aes(f1, f2)) +
    geom_point(alpha = 0.02, colour = "grey40") +                   
    geom_point(data = case_pts, aes(colour = case),                   
               alpha = 0.35, size = 1) +
    geom_point(data = case_ctr, aes(fill = case), shape = 21,   
               size = 3, colour = "black", stroke = 0.4, show.legend = FALSE) +
    ggrepel::geom_text_repel(data = case_ctr, aes(label = case),
                             size = 3, max.overlaps = Inf,
                             box.padding = 0.4, min.segment.length = 0,
                             seed = 42) +
    guides(colour = "none") +
    labs(x = "F1", y = "F2") +
    theme_minimal()


# library(mgcv)
# 
dd <- plt_scored %>%
    filter(!is.na(theta), !is.na(leader_first_year),
           !is.na(region), !is.na(polity_name)) %>%
    mutate(polity_name = factor(polity_name),
           region      = factor(region))

dd2 <- dd %>%
    group_by(polity_name) %>%
    filter(n() >= 5) %>%
    mutate(theta_c = theta - mean(theta)) %>%
    ungroup()

m_fe <- bam(theta_c ~ s(leader_first_year, k = 30), data = dd2, discrete = TRUE)
plot(m_fe, shade = TRUE)


m_reg <- bam(theta_c ~ region + s(leader_first_year, by = region, k = 25),
            data = dd2, discrete = TRUE, nthreads = 4)

plot(m_reg, select = 1, shade = TRUE)

plt_gemini %>% filter(polity_name == "Roman Republic") %>%
    count(cut(leader_first_year, seq(-510, -20, by = 50))) %>% print(n = 20)
```

## Lexical vs. IRT

```{r Lexical vs IRT}
# theta   : m1    (Version C, no tenure / sovereign)
# theta_B : mod_B (Version B, includes tenure_3_prediction_w3). lexical_lw3 and
#           theta_B share the tenure information, so their correlation is partly mechanical
fs_B <- fscores(mod_B, method = "EAP")

# Same sample as the lexical scale (known years, entry, exit). New object so the
# positional cor(plt_scored$theta, plt_gemini$...) below still lines up
lex_irt <- plt_scored %>%
    mutate(theta_B = fs_B[, 1]) %>%
    drop_na(leader_first_year, leader_last_year, entry_prediction, exit_prediction) %>%
    inner_join(lex %>% select(id, starts_with("lexical_"), matches("^lex[2-8]_")),
               by = "id")

stopifnot(nrow(lex_irt) == nrow(lex))

# 1. Spearman correlation (lexical is ordinal)
lex_irt %>%
    summarise(across(c(lexical_raw, lexical_w1, lexical_lw3),
                     list(theta   = ~cor(.x, theta,   method = "spearman", use = "complete.obs"),
                          theta_B = ~cor(.x, theta_B, method = "spearman", use = "complete.obs")))) %>%
    pivot_longer(everything(), names_to = c("lexical", "irt"),
                 names_pattern = "(lexical_[^_]+)_(.*)", values_to = "rho") %>%
    pivot_wider(names_from = irt, values_from = rho)

# 2. Mean theta by lexical level: should rise monotonically if the ladder is valid
lex_irt %>%
    group_by(lexical_raw) %>%
    summarise(n = n(),
              theta   = round(mean(theta,   na.rm = TRUE), 2),
              theta_B = round(mean(theta_B, na.rm = TRUE), 2))

lex_irt %>%
    pivot_longer(c(lexical_raw, lexical_lw3), names_to = "spec", values_to = "level") %>%
    ggplot(aes(factor(level), theta)) +
    geom_boxplot(outlier.alpha = 0.05) +
    facet_wrap(~spec) +
    labs(x = "Lexical level", y = "theta (Version C, no tenure)") +
    theme_minimal()

# 3. Tenure cap: within each uncapped level, theta of rows pushed down to level 1
#    vs. rows left alone. Capped rows with similar theta = cap disagrees with indicators
lex_irt %>%
    mutate(capped = lexical_raw > 1 & lexical_lw3 == 1) %>%
    group_by(lexical_raw, capped) %>%
    summarise(n = n(), theta = round(mean(theta, na.rm = TRUE), 2), .groups = "drop") %>%
    pivot_wider(names_from = capped, values_from = c(n, theta), names_prefix = "capped_")
```

## Check Indicator Order

```{r}
# graded_items <- names(type_B)[type_B == "graded"]
# 
# for (it in graded_items) {
#     a_it <- coef_B[it, "a"]
#     b <- coef_B[it, grep("^b[0-9]", colnames(coef_B))]
#     b <- b[!is.na(b)]
#     ok <- if (a_it >= 0) all(diff(b) > 0) else all(diff(b) < 0)
#     cat(sprintf("%-32s a = %6.3f   b = %-28s ordered? %s\n",
#                 it, a_it, paste(round(b, 3), collapse = "  "), ok))
# }
# 
# # fit_A <- itemfit(mod_A, fit_stats = "S_X2", na.rm = T)
# # fit_A %>% arrange(desc(RMSEA.S_X2)) %>% print()
# # 
# fit_B <- itemfit(mod_B, fit_stats = "S_X2", na.rm = T)
# fit_B %>% arrange(desc(RMSEA.S_X2)) %>% print()
```

## Local Dependence Problem (assembly; check_assembly; elections)

```{r}
# q3_full <- residuals(mod_A, type = "Q3", verbose = FALSE)
# q3_pairs <- as.data.frame(as.table(q3_full)) %>%
#     filter(as.character(Var1) < as.character(Var2)) %>%
#     rename(item1 = Var1, item2 = Var2, Q3 = Freq) %>%
#     arrange(desc(abs(Q3)))
# head(q3_pairs, 20)
# 
q3_full_B <- residuals(mod_B, type = "Q3", verbose = FALSE)
q3_pairs <- as.data.frame(as.table(q3_full_B)) %>%
    filter(as.character(Var1) < as.character(Var2)) %>%
    rename(item1 = Var1, item2 = Var2, Q3 = Freq) %>%
    mutate(Q3 = Q3 + 0.1) %>% # adjust negative expectation
    arrange(desc(abs(Q3)))
head(q3_pairs, 20)
```

```{r}
# plt_gemini %>%
#     count(sovereign_prediction, exit_4_prediction) %>%
#     group_by(sovereign_prediction) %>%
#     mutate(prop = round(n / sum(n), 3)) %>%
#     select(-n) %>%
#     tidyr::pivot_wider(names_from = sovereign_prediction,
#                        values_from = prop,
#                        names_prefix = "sovereign_")

cor(plt_scored$theta, plt_gemini$representation, use = "complete.obs", method = "spearman")
```

```{r}
# see if the second dimension shares similar trend with year
# mod2 <- mirt(dat_B, 2, itemtype = type_B, method = "EM", rotate = "oblimin")
# f <- fscores(mod2, method = "EAP")
# 
# cor(f[, 2], plt_gemini$leader_first_year, use = "complete.obs") 
# summary(mod2)  
# 
# # check symbolism labeling
# it_nosym <- unname(type_B)[items_B != "symbolism_prediction"]
# mod_nosym <- mirt(dat_B[, items_B != "symbolism_prediction"], 1,
#                   itemtype = it_nosym, method = "EM")
# plt_gemini %>%
#     mutate(th_ns = fscores(mod_nosym, method = "EAP")[, 1]) %>%
#     group_by(symbolism_prediction) %>%
#     summarise(n = n(), theta = mean(th_ns, na.rm = TRUE)) %>%
#     arrange(theta)
```

```{r}
plot(mod_B, type = "info")
plot(mod_B, type = "SE")

plot(mod_B, type = "infotrace", facet_items = TRUE)

th <- fscores(mod_B, method = "EAP", full.scores = TRUE, full.scores.SE = TRUE)

plt_gemini %>%
    mutate(theta = th[, "F1"], se = th[, "SE_F1"],
           era = cut(leader_first_year,
                     breaks = c(-Inf, 500, 1500, 1800, Inf),
                     labels = c("Ancient","Medieval","EarlyModern","Modern"))) %>%
    group_by(era) %>%
    summarise(n = n(), mean_theta = mean(theta), mean_se = mean(se))

```

## Theta

```{r}
# fs_A <- fscores(mod_A, method = "EAP", full.scores = TRUE, full.scores.SE = TRUE)
# fs_B <- fscores(mod_B, method = "EAP", full.scores = TRUE, full.scores.SE = TRUE)
# 
# # empirical reliability
# empirical_rxx(fs_A)
# empirical_rxx(fs_B)
# 
# plt_gemini <- plt_gemini %>%
#     mutate(theta_A = fs_A[, "F1"], theta_A_se = fs_A[, "SE_F1"],
#            theta_B = fs_B[, "F1"], theta_B_se = fs_B[, "SE_F1"])
# 
# cor(plt_gemini$theta_A, plt_gemini$theta_B, use = "complete.obs", method = "spearman")
# 
# # Simple sum
# plt_gemini <- plt_gemini %>%
#     mutate(simple_sum = rowSums(across(all_of(items_A)), na.rm = TRUE))
# 
# cor(plt_gemini$theta_A, plt_gemini$simple_sum, use = "complete.obs")
# 
# ggplot(plt_gemini, aes(x = theta_A, y = theta_B)) +
#     geom_hex(bins = 60) +
#     scale_fill_viridis_c(trans = "log10") +
#     labs(x = "theta (Version A: 19 items)",
#          y = "theta (Version B: checks summed)") +
#     theme_minimal()
```

# Bifactor

```{r}

```

## EFA

```{r}
# This does not converge
# mod_efa <- mirt(dat_A2, model = 3, itemtype = type_A2, verbose = FALSE)
# summary(mod_efa, rotate = "bifactorQ")
```
